---
title: "Lab_9"
author: "Mark Chafin"
date: "3/30/2021"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(dplyr)
library(tidyr)
library(ggplot2)
```

# 1
Download Missouri S&T's enrollment file “mst_enrollment2015.csv  download” and read it into R.
```{r}
df <- read.csv("mst_enrollment2015.csv")
```

# 2
Show all the unique countries of MST students. (1 point) How many countries are there? (1 point) 
```{r}
unique_countries <- unique(df$Country)
length(unique_countries)
# There are 89 unique countries
```

# 3
Then let's focus on domestic students and use dplyr::filter() to select students come from the US. (2 point)
```{r}
df <- df[-1]
df <- df %>% filter(Country == "United States")
```

# 4
Use tidyr:: gather to reshape the dataset by transforming it from wide format to long format, using “Year” as key and “Enroll” as value. (2 points)

```{r}
df <- df %>% gather(Year, Enroll ,c(X2011,X2012,X2013,X2014,X2015))
df
```
# 5
Use dplyr::group_by to aggregate enrollment data into state level, i.e. calculate total enrollment for each state. (2 points)

```{r}
df <- df %>% group_by(State) %>% summarise(Total_Enroll = sum(Enroll))

df
```
# 6
Since our task is to plot out-of-state enrollment, use dplyr::filter() to remove enrollment data of Missouri (in-state enrollment). (2 point)

```{r}
df <- df %>% filter(State != "Missouri")
```
# 7
Show the top 6 states from which the out-of-state students come and their corresponding enrollment number. (2 points)

```{r}
df <- arrange(df,-Total_Enroll)
head(df)
```
# 8
Use ggplot2 package to draw the MST's out-of-state enrollment on the US states map. (6 points)

```{r}
us_states <- map_data("state")
str(us_states)
map_state <- unique(us_states$region)
df$Region <- tolower(df$State)
setdiff(df$Region,map_state)
```

```{r, warning=FALSE}
gg <- ggplot()
gg <- gg + geom_map(data = us_states, map = us_states,
                    aes(x = long, y = lat, map_id = region),
                    fill="#ffffff", color="#ffffff", size=0.15)

gg


gg <- gg + geom_map(data=df, map = us_states,
                    aes(fill=Total_Enroll, map_id=Region),
                    color="#ffffff", size=0.15)
gg

```

```{r}
gg <- gg + scale_fill_continuous(low = "lightgreen", high= "darkgreen",
                                 guide="colorbar")
gg <- gg + labs(fill = "Enrollment",
                title = "US Enrollment",
                x = "", y="")

gg <- gg + theme(panel.border = element_blank())
gg <- gg + theme(panel.background = element_blank())
gg <- gg + theme(axis.ticks = element_blank())
gg <- gg + theme(axis.text = element_blank())

gg
```

# 9
Knit your R Markdown file into an HTML file and upload both R Markdown file and associated HTML file to Canvas (2pts).

```{r}

```

