---
title: "Working with Relational Database"
author: "Cui Zou"
date: "Sep 13, 2020"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage

In this example, we'll learn how to access a simple RDBMS, SQLite.

We use the RSQLite package. If you haven't installed the RSQLite package, please run the following command:

install.packages("RSQLite")

```{r}
# Clean the environment
rm(list = ls())
```


# 1. Connect to SQLite

First, we need to load RSQLite package. Then we call the dbConnect() function in RSQLite package to connect to a SQLite database.

```{r}
# Load RSQLite package
library("RSQLite")

# Specify the database file
dbfile <- "TelecomService.sqlite"

# Create a database connection
con <- dbConnect(dbDriver("SQLite"), dbname = dbfile)

```

# 2. List tables existing in database

```{r}
# Show existing tables
dbListTables(con)

```

# 3. Read Data from SQLite

After connection established, we can call dbGetQuery() function to get the result of SQL query.

```{r}
# Get call data
call_data <- dbGetQuery(con, "SELECT * FROM Call")

head(call_data)
summary(call_data)
str(call_data)

```

```{r}
# Get contract data
contract_data <- dbGetQuery(con, "SELECT * FROM Contract")

head(contract_data)
summary(contract_data)
str(contract_data)

```

```{r}
# Get both call and contract data
all_data <- dbGetQuery(con,
  "SELECT * FROM Call a, Contract b WHERE a.AreaCode = b.AreaCode AND a.Phone = b.Phone")

head(all_data)
summary(all_data)
str(all_data)

```

Drop the last two columns as they contain the same information as column 15 & 16
```{r}
all_data <- subset(all_data, select = -c(22, 23))
head(all_data)
```





