---
title: "Working with APIs"
author: "Cui Zou"
date: "Sep 13, 2020"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage

In this example, we'll learn how to access data via application programming interfaces (APIs) provided by Steam.

We need the rjson package. If you haven't installed the package, please run the following command:

install.packages("rjson")

Check the following website for the API specification: https://developer.valvesoftware.com/wiki/Steam_Web_API

In the example, we choose "GetNewsForApp (v0002)" API to get the latest news of a game specified by its appID.

```{r}
# Clean the environment
rm(list = ls())
```

Set the AppID of game "Grand Theft Auto V". Set the number of latest news we want to get.


```{r}
appid <- 271590 # Choose Grand Theft Auto V whose App ID is 271590
n_new <- 100 # Number of news to get
```


# 1. Get URL for the App

```{r}
# Generate url for the app
url <- paste("http://api.steampowered.com/ISteamNews/GetNewsForApp/v0002/?appid=",
             appid,
             "&count=", n_new, "&maxlength=300&format=json",
             sep = "")

# Check the url again
url
```

# 2. Get HTTP Content

```{r}
# Download HTTP response
content = RCurl::getURI(url)

# Print the first 800 characters
print(substr(content,1,800))

```

From the above, we can find that the HTTP return is in JSON format. Now let's convert it into a list structure for better data management in R.

```{r}
# Create a list
library("rjson")
list <- fromJSON(content)

# Show content in the list
head(list$appnews$newsitems)

```

Show the 1st news item.

```{r}
list$appnews$newsitems[[1]]
```

Show the 2nd news item.

```{r}
list$appnews$newsitems[[2]]
```


# 3. Organize Data in Data Frame

```{r}
# Create a data frame
# unlist function converts the list to vector with all the
# atomic components are being preserved.

df <- data.frame(matrix(unlist(list$appnews$newsitems), 
                        ncol = 11, 
                        nrow = n_new,
                        byrow=T)
                 )

str(df)

```

```{r}
head(df)
```

You'll notice that the column names do not make sense. Let's rename them.

```{r}
# Rename variables in the data frame
colnames(df) <- names(list$appnews$newsitems[[1]])

# show head of the data frame.
head(df)

```

You'll notice that the date is shown as numbers (in strings actually). Let's convert it into a readable data type. Here we can use POSIXlt data type to represent data time. To learn more about this data type, type "?POSIXlt" in console window.

```{r}
# Add variable date2 to show the readable date time
df$date <- as.POSIXlt(as.numeric(as.character(df$date)),origin = "1970-01-01")

# Show head again
head(df)

```

To learn more: Below shows how to convert between numbers and data time.
 
```{r}
now <- Sys.time()
print(now)
```

```{r}
n <- as.numeric(now)
print(n)
```

To convert the number into data time.

```{r}
as.POSIXlt(n,origin = "1970-01-01")
```

