---
title: "Lab_8"
author: "Mark Chafin"
date: "3/15/2021"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```
# 1 
Download the cleansed weblog analytics data file “weblog_clean.jsonl” and read it into R (2pts). Note: The data file is in the format of JSON Lines. You should load “jsonlite” package and call stream_in() function to stream JSON input:
library(jsonlite)
weblog <- stream_in(file("weblog_clean.jsonlines"))

```{r}
library(dplyr)
library(jsonlite)
weblog <- stream_in(file("weblog_clean.jsonlines"))
```

# 2
Show the structure of the dataset.

```{r}
str(weblog)
```

# 3
Change the data type of “request.datetime” as POSIXct by using the as.POSIXct() function (2 pts).

```{r}
weblog$request.datetime <- as.POSIXct(weblog$request.datetime, format = "%Y-%m-%d %H:%M:%S")
```

# 4
Show summary statistics of the dataset.
```{r}
summary(weblog)
```

# 5
Get the frequency distribution of weekday by using table() function (1 pt). Which day has the largest number of visits (1 pt)?
```{r}
weekday.freq <- table(weblog$weekday)
sort(weekday.freq)
print("Has the largest number of visits")
weekday.freq %>% which.max() 
```

# 6
Cross tabulate weekday and request.uri (1 pt)(refer to the two-way frequency table on slide 17 of Chapter 5). How many visits of the “faq.html” page on Fridays (1 pt)? 
```{r}
cross_tabulate <- table(weblog$weekday, weblog$request.uri)
cross_tabulate 
print("There are 4407 visits on the faq.html page on Friday")
```

# 7
Draw a bar plot to show the distribution of visits by weekday (2 pts).
```{r}
weblog.barplot <- table(weblog$weekday)
barplot(weblog.barplot, col = rainbow(20), cex.names = .8)
```

# 8
Draw a stacked bar plot to show the distribution of visits by request.uri and request.method (use request.method to group visits) (2 pts).
```{r}
weblog.stacked <- table(weblog$request.method, weblog$request.uri)
barplot(weblog.stacked,
        main = "Distributuion of Visits",
        col = rainbow(3),
        legend = rownames(weblog.stacked))
```

# 9
Draw a pie chart to present the distribution of visits by request.uri and show the count of visits for each URI (2 pts).
```{r}
weblog.piechart <- table(weblog$request.uri)
labels <- paste(names(weblog.piechart),"\n", weblog.piechart, sep = "")
pie(weblog.piechart, labels = labels, col = rainbow(5))
```

# 10
Draw a pie chart to present the distribution of visits by request.uri and how the percentage of visits for each URI (2 pts).
```{r}
piechart <- round(weblog.piechart/sum(weblog.piechart)*100, digits = 1)
label2 <- paste(names(weblog.piechart), " visits\n", paste(piechart,"%"), sep="")
pie(piechart,
    labels = label2,
    col=rainbow(length(label2)))
```

# 11
Calculate 5-number summary for reponse.size by using quantile() function (2 pts).
```{r}
quantile(weblog$response.size)
```