---
title: "Cleansing and Manipulating Dataset"
author: "Cui Zou"
date: "Oct 3, 2020"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage

Cleansing and manipulating data is an essential (usually the most time-consuming) part of a data analytics project.
A properly prepared dataset is the prerequisite of statistical modeling, prediction, and inference.

In this section, we'll learn how to cleanse, transform, manipulate dataset.

```{r}
# Clean the environment
rm(list = ls())
```

# 0. Forward Pipe Operator

The forward pipe operator is supported by magrittr package. The syntax is "lhs %>% rhs". In order to use the forward pipe operator, you need to load the magrittr package. Or you can load the dplyr package which will also load the magrittr package.

To learn more about forward pipe operator, type "?magrittr::`%>%`" in console window.

```{r}
# Load magrittr package
library(magrittr)
# The data give the speed of cars and the distances taken to stop. 
# Note that the data were recorded in the 1920s.

speed <- cars$speed

print(speed)

```

```{r}
# Find the max speed
speed %>% max

```

```{r}
# List unique speed values
speed %>% unique

```

```{r}
# List unique speed values and sort them by descending order
speed %>% unique %>% sort(decreasing = TRUE)

```

```{r}
# List top 5 speed values and sort them by descending order
speed %>% unique %>% sort(decreasing = TRUE) %>% head(5)

```

When the lhs is not the first argument in rhs call, we can use the dot place-holder.

```{r}
# Use dot place-holder
"Ceci n'est pas une pipe" %>% gsub("une", "un", .)
```

Understand below syntax:

```{r}
sample(1:10) %>% paste0(LETTERS[.])
```

What if you want to put number after letter?

```{r}
sample(1:10) %>% paste0(LETTERS[.], .) 
```


# 1. Create, Recode, and Rename Variables

Let's create a data frame first, then use the data frame to show how to create, recode, and rename variables.

```{r}
patient <- data.frame(id = c("A01","A02","A03"),
                      first.name = c("Mike","Emily","Hannah"),
                      last.name = c("Smith","Johson", "Williams"),
                      age = c(26,20,24),
                      mass = c(150, 120, 110),
                      height = c(70,68,67))
print(patient)
```


## 1.1. Create Variables

### 1.1.1. Use index ($ operator)

```{r}
# Copy the original dataset
patient2 <- patient 
# Add new variable full.name
patient2$full.name <- paste(patient2$first.name,
                            patient2$last.name,
                            sep = " ")
# Add new variable bmi
patient2$bmi <- 703*patient2$mass/(patient2$height^2)

patient2 %>% print

```


### 1.1.2. Use transform() function

```{r}
# Use transform()
patient3 <- transform(patient,
                      full.name = paste(first.name,last.name,sep = " "),
                      bmi = 703*mass/(height^2))
patient3 %>% print
```

### 1.1.3. Use dplyr::mutate() function

```{r}
## Use dplyr::mutate() ##
patient4 <- dplyr::mutate(patient,
                          full.name = paste(first.name,last.name,sep = " "),
                          bmi = 703*mass/(height^2))
patient4 %>% print

```

## 1.2. Recode Variables

Recode the values of variables usually involves applying conditions.

```{r}
# Recode Variables
patient4$bmi.category[patient4$bmi>=16 & patient4$bmi<18.5] <- "Underweight"
patient4$bmi.category[patient4$bmi>=18.5 & patient4$bmi<25] <- "Normal"

patient4$age.group = ifelse(patient4$age>=25,"Older","Younger")

patient4 %>% print

```

```{r}
# Recode the ID
patient4$id <- dplyr::recode(patient4$id,A01 = "B01", A02 = "B02", A03 = "B03")

# Recode the age
patient4$age <- dplyr::recode(patient4$age,`26` = 27L,`20` = 21L,`24` = 25L)

patient4 %>% print

```

## 1.3. Rename Variables

```{r}
# Rename bmi.categrory as health.status
names(patient4)[names(patient4) == "bmi.category"] <- "health.status"

# Rename mass as weight
patient4 <- dplyr::rename(patient4,weight = mass)

patient4 %>% print

```


# 2. Convert Data

Here, we read the dataset we collected from https://nrf.com/2015/top100-table.

```{r}
df <- read.csv("top100retailers2015.csv")
# Show the structure of the dataset
str(df)

```

```{r}
summary(df)
```

We can notice that all variables except the first two are factor data (i.e., nominal scale). As a result, we cannot get correct summary statistics such as min, max etc. for nominal data.

We need to convert these factors as numeric variables. The following example shows the conversion of the column named "RetailSales2014". Other columns/variables can be converted in the similar way.

```{r}
head(df$RetailSales2014)
```


```{r}

df$RetailSales2014 <- df$RetailSales2014 %>%
  gsub(",","",.) %>%
  gsub("$","",.,fixed = TRUE) %>%
  as.character() %>%
  as.numeric()

# Show the structure of the new variable
str(df$RetailSales2014)

```

```{r}
# Get summary statistics of the new variable
summary(df$RetailSales2014)
```


# 3. Sort

Load the built-in dataset mtcars. The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).


```{r}
# Load dataset
data(mtcars)

# Show the strucure
str(mtcars)
```


## 3.1. Use order() function in base R

```{r}
# Sort by mpg (ascending)
mtcars[order(mtcars$mpg),]

```
```{r}
# Sort by cyl (ascending) and hp (ascending)
mtcars[order(mtcars$cyl, mtcars$hp),]

```

```{r}
# Sort by cyl (ascending) and hp (descending)
mtcars[order(mtcars$cyl, -mtcars$hp),]
```


## 3.2. Use arrange() function in dplyr package

```{r}
# Sort by mpg (ascending)
mtcars %>% 
  dplyr::arrange(mpg) %>%
  head

```

```{r}
# We don't know which cars have lowest mpg
# Convert row names to an explicit variable
cars <- tibble::rownames_to_column(mtcars, "brand")
cars

```


```{r}
# Sort by cyl (ascending) and hp (ascending)

cars %>% 
  dplyr::arrange(cyl,hp) %>%
  head

```


```{r}
# Sort by cyl (ascending) and hp (descending)

cars %>% 
  dplyr::arrange(cyl,desc(hp)) %>%
  head

```


# 4. Subset

We can use subset methods to select certain columns or certain observations.

## 4.1. Use base R features

### 4.1.1. Use index

```{r}
# Use index to select variables
car1 <- mtcars[c("mpg","wt")]  # Select miles per gallon and weight
names(car1)

```

```{r}
car2 <- mtcars[c(1,6)]  # Select 1st and 6th variables, i.e. mpg and wt
names(car2)

```

```{r}
car3 <- mtcars[c(1:5,10)]  # Select 1st through 5th and 10th variables
names(car3)

```

```{r}
car4 <- mtcars[c(-3,-5)]  # Exclude the 3rd and 5th variables
names(car4)

```


### 4.1.2. Use which() function

```{r}
# Use index and which() to select observations
car5 <- mtcars[1:4,]  # First 4 observations
car5
```

```{r}
car6 <- mtcars[which(mtcars$mpg >= 25),]  # Select cars whose mpg >= 25
car6

```

```{r}
car7 <- mtcars[which(mtcars$mpg >= 25 & mtcars$gear == 4),]  # Select cars whose mpg >= 25 and gear == 4
car7

```

```{r}
# Use index and which() to select both observations and variables
car8 <- mtcars[which(mtcars$mpg >= 25),c("mpg", "wt")]  # Select mpg and wt variables with mpg >= 25
car8
```


### 4.1.3. Use subset() function

```{r}
# Use subset() function to select both observations and variables
car11 <- subset(mtcars, select = c("mpg","wt"))  # Select mpg and wt variables
names(car11)

```

```{r}
car12 <- subset(mtcars, select = c(1,6))
names(car12)

```

```{r}
car13 <- subset(mtcars, select = mpg:wt)  # Select all variables between mpg and wt
names(car13)

```

```{r}
car14 <- subset(mtcars,mpg >= 25,select = c(mpg,wt)) # Select mpg and wt variables with mpg >= 25
car14

```

```{r}
car15 <- subset(mtcars,cyl == 6,select = c(1:6)) # Select 1st through 6th variables with cyl == 6
car15
```


## 4.2. Use dplyr package

```{r}
## Load dplyr package
library(dplyr)
```


### 4.2.1. Use dplyr::select() to select variables

```{r}
car21 <- select(cars,c(brand,mpg,wt))  # Select mpg and wt variables
names(car21)

```

```{r}
car22 <- select(cars,-c(mpg,wt))  # Select variables except mpg and wt
names(car22)

```

```{r}
car23 <- select(cars,contains("p"))  # Select variables containing "p"
names(car23)

```


### 4.2.2. Use dplyr::filter() to select observations

```{r}
filter(cars,mpg >= 25)  # Select cars with mpg >= 25

```

```{r}
filter(cars,mpg >= 25 & gear == 4)  # Select cars with mpg >= 25 and gear == 4

```

We can combine dplyr::select() and dplyr::filter().

```{r}
# Select mpg and wt variables with mpg >= 25 
cars %>% 
  filter(mpg >= 25) %>%
  select(c(brand,mpg,wt))
```

```{r}
# Another way
cars %>% 
  select(c(brand,mpg,wt)) %>%
  filter(mpg >= 25)

```


# 5. Merge

## 5.1. Use merge() function

```{r}
enrollment <- data.frame(courseId = "IST3420",
                         studentId = c("201","202","501"))

student <- data.frame(studentId = c("201","202","203"),
                      name = c("Mike","Emily","Hannah"),
                      department = c("IST","BUS","IST"))

print(enrollment)

```

```{r}
print(student)

```

```{r}
# Inner join
merge(x = enrollment, y = student, by = "studentId")

```

```{r}
# Left outer join
merge(x = enrollment, y = student, by = "studentId", all.x = TRUE)

```

```{r}
# Right outer join
merge(x = enrollment, y = student, by = "studentId", all.y = TRUE)

```

```{r}
# Full outer join
merge(x = enrollment, y = student, by = "studentId", all = TRUE)
```


## 5.2. Use join() function in plyr package

```{r}
# Load plyr package
library("plyr")

```

```{r}
print(enrollment)

```

```{r}
print(student)

```

```{r}
# Inner join
join(enrollment,student,by = "studentId", type = "inner")

```

```{r}
# Left outer join
join(enrollment,student,by = "studentId", type = "left")

```

```{r}
# Right outer join
join(enrollment,student,by = "studentId", type = "right")

```

```{r}
# Full outer join
join(enrollment,student,by = "studentId", type = "full")

```


## 5.3. Use dplyr package

The dplyr package provides 4 functions:

- inner_join()
- left_join()
- right_join()
- full_join()

```{r}
# Load dplyr package
library("dplyr")

```

```{r}
# Inner join
inner_join(enrollment,student,by = "studentId") # This may not work (depends on the version of dplyr)

```

```{r}
# Show the structure of data frames
str(enrollment)

```

```{r}
str(student)

```

```{r}
# Change vectors from factor type to character type
enrollment$courseId <- as.character(enrollment$courseId)
enrollment$studentId <- as.character(enrollment$studentId)
student$studentId <- as.character(student$studentId)
student$name <- as.character(student$name)
student$department <- as.character(student$department)

```

```{r}
str(enrollment)

```

```{r}
str(student)

```

```{r}
inner_join(enrollment,student,by = "studentId") # Now it works

```

```{r}
# Left outer join
left_join(enrollment,student,by = "studentId")

```

```{r}
# Right outer join
right_join(enrollment,student,by = "studentId")

```

```{r}
# Full outer join
full_join(enrollment,student,by = "studentId")
```


# 6. Aggregate

## 6.1. Use Base R

In base R, we can call aggregate() function to group data and calculate summary statistics.

```{r}
attach(mtcars)

# Use aggregate() function
# For cars with different combination of cyl & vs, get mean value of other variables
# and remove NA
mtcars %>% aggregate(by=list(cyl,vs),FUN=mean, na.rm=TRUE)

```


## 6.2. dplyr Package

In dplyr package, we can use:

- group_by(): to group data
- summarize(): to calculate summary statistics

```{r}
# Use dplyr::group_by() and dplyr::summarize() functions
mtcars %>% 
  group_by(cyl,vs) %>%
  summarize(mean(mpg),mean(disp),mean(hp),mean(drat),mean(wt),mean(qsec),mean(am),mean(gear),mean(carb))

```

**Wait, the above aggregation does not work! We expect to get five observations of the aggregated dataset. Now we only get one. Why?**


The reason is that we load plyr package in section 5.2. The plyr package has a function named "summarize". In the above R chunk, actually the function summarize() is the one in plyr package rather than the one in dplyr package. 

To solve the problem, we can simply provide package name for the summarize() function. Another way is to unload plyr package (executing statement "detach(package:plyr)") and then execute the above R chunk. In the following, we choice the first solution.

```{r}
# Use dplyr::group_by() and dplyr::summarize() functions
mtcars %>% 
  group_by(cyl,vs) %>%
  dplyr::summarize(mean(mpg),mean(disp),mean(hp),mean(drat),mean(wt),mean(qsec),mean(am),mean(gear),mean(carb))

```

```{r}
# Aggregate by cyl and vs, show mean of mpg for each group
mtcars %>% 
  group_by(cyl,vs) %>% 
  dplyr::summarize(mean(mpg))

```

```{r}
# Aggregate by cyl and vs, show maximum mpg for each group
mtcars %>% 
  group_by(cyl,vs) %>% 
  dplyr::summarize(max(mpg))
```


# 7. Reshape

We can use tidyr package to change the layout of a dataset. There are two functions to call:

- gather(data, key, value, .): to gather columns into rows
- spread(data, key, value): to spread rows into columns

```{r}
# Load tidyr package
library(tidyr)

```

## 7.1. Convert Wide Format to Long Format

```{r}
# A student score dataset in wide format
score_wide <- data.frame(
  name = c("Mike","Sarah","Helen"),
  IST3420 = c(80,85,90),
  IST1750 = c(78,81,91),
  IST3423 = c(90,93,88)
)
print(score_wide)

```

```{r}
## Gather columns into key-value pairs
score_long <- score_wide %>% 
  gather(course,score,IST3420, IST1750,IST3423)
print(score_long)

```

```{r}
# Another way to specify columns to gather
score_wide %>% 
  gather(course,score,c(IST3420, IST1750,IST3423))

```

```{r}
# Still another way to specify columns to gather
score_wide %>% gather(course,score,-name)

```

## 7.2. Convert Long Format to Wide Format

```{r}
## Print the long format
print(score_long)

```
```{r}
## Split a single character column into multiple columns
score_long %>% 
  spread(course,score)
```

