---
title: "R Notebook"
author: "markchafin"
date: "4/19/21"
output:
  html_document:
    df_print: paged
---

# 1
    Show the structure and head of the mtcars dataset. We focus on the following two variables:
mpg: miles per gallon
am: transmission (0 = automatic, 1 = manual)
```{r}
rm(list = ls())
library(dplyr)
library(ggplot2)

str(mtcars)
head(mtcars)
df <- select(mtcars, 1, 9)

mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c("Automatic", "Manual")
```

# 2
We want to visualize the difference of mpg between two types of transmission. Draw a boxplot of mpg grouped by transmission type (two boxplots on the same figure, each representing one transmission type) (2pts).
```{r}

ggplot(mtcars, aes(x=factor(am),y=mpg,fill=factor(am)))+
  geom_boxplot(notch=F)+
  scale_x_discrete("Transmission")+
  scale_y_continuous("Miles per Gallon")+
  ggtitle("MPG by Transmission Type")
```
# 3
What conclusion could you get from the above boxplot (2pts)?
Manual transmissions have a higher fuel efficiency
# 4
Now we want to use hypothesis testing method to test if manual transmission cars have better miles per gallon than automatic transmission cars. Specify your null and alternative hypothesis (4pts).

H0 mean(manualdata$mpg) < mean(autodata$mpg) 
Ha mean(manualdata$mpg) > mean(autodata$mpg) 

```{r}
autoData <- mtcars[mtcars$am == "Automatic",]
manualData <- mtcars[mtcars$am == "Manual",]
ttest <- t.test(autoData$mpg, manualData$mpg, alternative = "less", mu = 0)
ttest


autoData <- mtcars[mtcars$am == "Automatic",]
manualData <- mtcars[mtcars$am == "Manual",]
ttest <- t.test(autoData$mpg, manualData$mpg, alternative = "greater", mu = 0)
ttest



t.test(mpg~am, data = mtcars)

t.test(autoData$mpg, data = autoData)
t.test(manualData$mpg, data = manualData)


```
# 5
What is the appropriate hypothesis testing method: one-sample or two-sample (2pts)? Paired or unpaired (2pts)? Why (2pts)? Two-sample, unpaired. the two-sample is required to show the correlation between mpg and transmission (am) and you can't do a paired test unless you clean the data and make the sample sizes the same, since the two datasets are not equal.
# 6
Use the appropriate method from step 5 to test the hypothesis (2pts).
```{r}
t.test(mpg~am, data = mtcars, alternative = "less")
```

# 7
Explain the hypothesis result (2pts).
My hypothesis results show the manual cars have a higher mpg than automatic cars. This data is from 1974 Motor Trend US magazine. Todays data could be different.

# 8
Knit your R Markdown file into an HTML file and upload both R Markdown file and associated HTML file to Canvas (2pts).