---
title: "Working with HTML"
author: "Cui Zou"
date: "Sep 13, 2020"
output:
  html_document:
    df_print: paged
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage

In this example, we'll learn how to scrape data table from the following Wikipedia web page: 

https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population

We'll use the XML and httr packages. If you haven't installed them, please run the following command:

install.packages("XML")
install.packages("httr")

The R Markdown has been updated on 9/13/2020.

```{r}
# Clean the environment
rm(list = ls())
```

```{r}
# Load packages
library("XML")
library("httr")
# Or you can use require() to load needed packages
```


# 1. Download HTML Content from Web

First, specify the URL of the web page where we want to scrape data.

```{r}
# Specify URL
url <- "https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population"

print(url)
```

Call the GET() function in the httr package to download the content of the web page.

```{r}
# Download the content of the URL
url_content <- GET(url)

print(substr(url_content,1,800))

```

# 2. Parse the HTML Content

```{r}
# Parse the HTML/XML content to generate an R structure representing the HTML/XML tree
doc <- htmlParse(url_content)

tables <- readHTMLTable(doc)

str(tables)
```


```{r}
# Convert the 1st element of the list to data frame
pop_df <- data.frame(tables[1])

attributes(pop_df)

```

# 3. Refine the Dataframe

Show the structure of the dataframe.

```{r}
str(pop_df)
```

Show the head of pop_df

```{r}
head(pop_df)
```


Rename the column names of the dataframe.

```{r}
colnames(pop_df) <- c("Rank","Country/Territory","Population","Date","% of World Population","Source")

is.factor(pop_df$Population)

# Convert factors into numbers for Population column
pop_df$Population <- as.numeric(gsub(",","",pop_df$Population))  # Use gsub() to remove all commas in Population

str(pop_df)

```

Delete 1st row of pop_df

```{r}
pop_df_new <- pop_df[-c(1),]
head(pop_df_new)
```

Change row names

```{r}
row.names(pop_df_new) <- 1:nrow(pop_df_new)
head(pop_df_new)
```



Show the top 10 countries/territories.

```{r}
top10 <- head(pop_df_new, n = 10L)

plot(top10$Population, type = "l")

```
