---
title: "Working with XML and JSON"
author: "Cui Zou"
date: "Sep 13, 2020"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage

In this example, we'll learn how to access data stored in XML and JSON.

We use the XML package. If you haven't installed the XML package, please run the following command:

install.packages("XML")

```{r}
# Clean the environment
rm(list = ls())

# Load packages
library("XML")
library("RCurl")

```


# 1. Access XML

## 1.1. Read XML

First, we need to load XML package. Then we can call the xmlParse() function in XML package to parse and XML file to generate an R structure.

```{r}
# Load XML package
library("XML")

# Parse XML file to generate an R structure
doc <- xmlParse(file = "Books.xml")

# Show the content of doc
print(doc)

```

```{r}
# Get the root node
top <- xmlRoot(doc)

# Show the content of top
print(top)

```

## 1.2. Explore XML

Now, let's explore the root node of the XML document.

```{r}
# Show tage name of the root node
xmlName(top)
```

```{r}
# Show child nodes of the root node
names(top)
```

```{r}
# Show child nodes of the 1st book
names(top[[1]]) 
```

```{r}
# Show title of the 1st book
top[[1]][["title"]]

```

```{r}
# Show year of the 1st book
top[[1]][["year"]] 
```

```{r}
# Show price of the 1st book
top[[1]][["price"]] 
```

## 1.3. Use XPath to Parse XML

We can use xpathSApply() to extract information by XPath expression. The xpathSApply() is a simplified version of xpathApply() function.

First, let's parse the title information from the XML object and store the information in a vector.

```{r}
b_title <- xpathSApply(doc,"//bookstore/book/title", xmlValue)

print(b_title)
```

Use the same method to parse other information.

```{r}
b_category <- xpathSApply(doc,"//bookstore/book/@category")
b_author <- xpathSApply(doc,"//bookstore/book/author", xmlValue)
b_year <- xpathSApply(doc,"//bookstore/book/year", xmlValue)
b_price <- xpathSApply(doc,"//bookstore/book/price", xmlValue)
```

## 1.4. Organize Information in Data Frame

Now we have obtained all information stored in the XML object. Let's create a data frame to organize the structured information.

```{r}
# Generate a book data frame
book_df <- data.frame(b_title,b_category,b_author,b_year,b_price)
colnames(book_df) <- c("Title","Category","Author","Year","Price")
print(book_df)
```


# 2. Access JSON

Compared with XML, JSON is usually easier to access in R.

We use the jsonlite package. If you haven't installed the package, please run the following command:

install.packages("jsonlite")

```{r}
# Load jsonlite package
library(jsonlite)

```

## 2.1. Convert between Data Frame and JSON

We can convert a data frame to JSON.

```{r}
# data.frames must be converted into a list before converting into JSON
j_book <- toJSON(as.list(book_df))
print(j_book)

```

Or conver JSON back to data frame.

```{r}
# Convert JSON to data frame
book_df2 <- data.frame(fromJSON(j_book))
print(book_df2)
```

## 2.2. Read or Write JSON Files

If we want to write the JSON object to a file, we can use the write_lines() function in readr package.

```{r}
library(readr)

write_lines(j_book,"books.json")
```

