---
title: "factor2numeric"
author: "Cui Zou"
date: "9/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Why we can't use as.numeric() to convert factor directly into numeric?

```{r}
library(magrittr)

# convert a numeric vector into factor
num <- factor(c(9,8,10,8,9))
str(num)

# intend to convert factor into numeric, but only return level of each element
num1 <- as.numeric(num)
str(num1)
num1

# the mothod above is not working, let's change it into character first
num2 <- as.character(num)
str(num2)

# use as.numeric(as.character()) to convert factor to numeric
num3 <- num %>% 
  as.character %>%
  as.numeric
str(num3)
num3
```

## 2. What if we convert all factor variables in a data frame to numeric variables?

mutate_if(df, is.factor, ~ as.numeric(as.character(.x)))

Or

mutate_if(df, is.factor, ~ as.numeric(levels(.x))[.x])

** The latter is faster when you work with large data frame.**

## 3. How to avoid unnecessary factors when read in data?

Use the statement stringsAsFactors = FALSE
e.g. df <- read.table(<path_to_your_data>, sep = "\t", stringsAsFactors = FALSE)

source: https://community.rstudio.com/t/how-to-convert-factor-data-to-numeric-for-all-dataset/17626



