---
title: "Manipulating Strings"
author: "Cui Zou"
date: "Oc 3, 2020"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage

Cleansing and manipulating data is an essential (usually the most time-consuming) part of a data analytics project. A properly prepared dataset is the prerequisite of statistical modeling, prediction, and inference. 

In this section, we'll learn how to manipulate strings.

```{r}
# Clean the environment
rm(list = ls())
```

In many cases, you have to deal with textural data. There are many useful string manipulation functions in R. Below is a list of frequently used ones.

# 1. General Functions

- tolower(): translate to lower case
- toupper(): translate to upper case
- nchar(): count the number of characters
- trimws(): trim whitespace [ \\t\\r\\n]
- stringi::stri_reverse(): reverse a string

As an example, we use the vector of row names of the mtcars dataset to show how to manipulate a vector of strings.

```{r}
car <- rownames(mtcars)

print(car)
```

```{r}
# Translate to lower case
tolower(car)
```

```{r}
# Translate to upper case
toupper(car)

```

To count the number of characters in a string, do NOT use length(). length() returns the length of the vector containing the string. Instead, use nchar() function.

```{r}
# Count the number of characters
nchar(car)

```

```{r}
# The length of the vector
length(car)

```

To trim whitespace [ \\t\\r\\n] (space, tab, newline, carriage return), we can use trimws() function.

```{r}
trimws("\n  Hello R! \t \r \n   ")

```

To reverse strings, we can use the stri_reverse() function in stringi package. To call a function not available in base R system, we can apply two methods: 

- Method 1: Directly calling the function without loading the package by using "::";
- Method 2: Load the package first, then call the function.


```{r}
# Directly use the function without loading the package
stringi::stri_reverse("abcdef")  

```

```{r}
# Or use the following statements
require(stringi)
stri_reverse("abcdef")
```


# 2. Pattern Matching and Replacement

Regular expression is a pattern that describes a set of strings. Regular expression is a powerful technique for string matching and processing.

To learn more about regular expressions in R, type ?regex in console window.

- grep(): detect patterns
- regexpr(): locate patterns (find starting position and length of first match)
- sub(): replace the first match
- gsub(): replace all matches

For example, match a substring of 4 consecutive digits from the string "1/Jun/2016 2:35 PM". That is, to find the year information from the string. Here, we use regular expression "[0-9]{4}".

```{r}
# Match a pattern in a string
pos1 <- regexpr("[0-9]{4}","1/Jun/2016 2:35 PM")
pos1

```

```{r}
# Position of the 1st match in the string
pos1[[1]]
```

Another way is to use regex "[[:digit:]]{4}".

```{r}
# Match a pattern in a string
pos1 <- regexpr("[[:digit:]]{4}","1/Jun/2016 2:35 PM")
pos1

```

In the next example, we match all substrings of digit 2. We simply use "2" as the pattern to match.

```{r}
# Positions of all matches in the string
pos2 <- gregexpr("2","1/Jun/2016 2:35 PM")
pos2
```

```{r}
# Position of the 1st match
pos2[[1]][1]

```

```{r}
# Position of the 2nd match
pos2[[1]][2]

```

Another function to call is grep(). Those functions are similar. They differ in the format of and amount of detail in the results.

```{r}
# Show car vector
print(car)

```


```{r}
# The index of matched string in the vector 
grep("Merc",car)

```


To replace matches, we can use sub() or gsub() function.

```{r}
# sub() replaces the first match
sub("a","1","abcabcabc")

```

```{r}
# gsub() replaces all matches
gsub("a","1","abcabcabc") 

```

```{r}
# Remove all digits by using regex
gsub("[[:digit:]]","", "a1b2c3ef4gh55") 

```

# 3. Substrings

- substr(x, start, stop)
- substring(text, first, last = 1000000L)
- substr vs substring?  substring has a default value for the last option: last = 1000000L while substr doesn't have a default value for the stop option.

```{r}
substr("abcdef", 2, 5)

```

```{r}
substring("abcdef", 1, 1:6)

```

```{r}
substring("abcdef", 1:6, 1:6)
```


# 4. Split Strings

- strsplit(x, split, fixed = FALSE, perl = FALSE, useBytes = FALSE)

```{r}
strsplit("abcdef", NULL)  # Format: strsplit(x, split)

```

```{r}
strsplit("a.b.c.d.e.f", "[.]")  # split argument is a regular expression

```

```{r}
strsplit("a.b.c.d.e.f", ".", fixed = TRUE) # Use exact matching

```

```{r}
strsplit("12 23   14 21 56 78 99  "," ")

```

```{r}
strsplit("12 23   14 21 56 78 99  ","[[:blank:]]+") # Match blank characters one or more times

```

From above examples, you can find that the result of strsplit() function call is a list. To get a vector, we can unlist the result.

```{r}
unlist(strsplit("12 23   14 21 56 78 99  ","[[:blank:]]+"))  # To get a vector
```


# 5. Concatenate

To concatenate string objects, we can call paste() or paste0(). The function paste0(...) is equivalent to paste(..., sep = ""), slightly more efficiently.

-paste()
-paste0()

```{r}
paste("abc","123",sep = "")

```

```{r}
paste("abc","123",sep = ",")

```

```{r}
paste0("abc","123")

```

```{r}
nth <- paste0(1:12, c("st", "nd", "rd", rep("th", 9)))
nth

```

```{r}
month.abb

```

We can specify the "collapse" parameter to collapse the vector into a single string element with a character string to separate the parts.

```{r}
v1 <- paste(month.abb,nth,sep = ": ",collapse = "; ")
nchar(v1)
length(v1)
```
