---
title: "Lab_4"
author: "Mark Chafin"
date: "2/7/2021"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Import libraries
```{r}
library(XML)
library(httr)
library(dplyr)
```
# 1. Read the HTML Table
# 2. Store the HTML table as a df in Facebook_df
```{r}
url <- GET("https://finance.yahoo.com/quote/FB/history?p=FB")
facebook_df <- as.data.frame(readHTMLTable(htmlParse(url)))
```
# 3. Show the attributes and structure of the df
```{r}
attributes(facebook_df)
str(facebook_df)
```
# 4. List original column names
```{r}
colnames(facebook_df)
```
# 5. Rename columns
```{r}
facebook_df <- facebook_df %>% 
  rename(Date = NULL.Date,
       Open = NULL.Open,
       High = NULL.High,
       Low = NULL.Low,
       Close = NULL.Close.,
       "Adj Close" = NULL.Adj.Close..,
       Volume = NULL.Volume)
colnames(facebook_df)
```
# 6. Show summary statistics of the df
```{r}
summary(facebook_df)
```
# 7. Do you get the min, median, mean, and max values of the facebook_df variables (1pts)? No
Why (1pts)? The function `summary()` retrieves the summary statistics of the df called but since the values in the df are not numbers `summary()` cannot return `min` `max` `median` `mean`
```{r}

```

# 8. Save the data frame as a CSV file “facebook_prices.csv”
```{r}
write.csv(facebook_df, file = "facebook_prices.csv", row.names = F)
```

