---
title: "Lab_6"
author: "Mark Chafin"
date: "3/1/2021"
output:
  
  html_document:
  df_print: paged
  toc: yes
  toc_depth: '3'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(dplyr)
library(tidyr)
```

# 1 Download “top100retailers2015.csv” and read the csv file into R.
```{r}
df <- read.csv("top100retailers2015.csv")

```
# 2 Get its structure and summary statistics.
```{r}
str(df)
summary(df)
```

# 3 Use regular expressions and forward pipe operators (magrittr::%>%) to remove special characters (e.g. $ % ,) from the following columns, then convert those columns into numeric data [use as.numeric(as.character())]
```{r}

for (i in 5:10) {
   df[,i] <- df[,i] %>%
    gsub("[[:punct:]]","",.) %>% 
    as.numeric(as.character())
}
head(df)
str(df)
```

# 4 Create “City” and “State” columns from “Headquarter” [use regexpr() and substr(), or tidyr::separate()] 
```{r}
df <- separate(df, Headquarter, sep = ",", into =c("City", "State"))
summary(df)

```

# 5 Show summary statistics of the new dataset and remove variable X
```{r}
summary(df)
df <- subset(df, select = -X)
```

# 6 Sort column RetailSales2014 in descending order and show the head 
```{r}
df <- df %>% 
  arrange(desc(RetailSales2014))
head(df)
```

# 7 Save the cleansed dataset as top100retailers_clean.csv
```{r}
write.csv(df, file = "top100retailers_clean.csv")
```















