---
title: "Data Exploration - Hypothesis Testing"
author: "Cui Zou"
date: "Oct 11, 2019"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage


```{r}
# Clean the environment
rm(list = ls())
```

There are two methods t-test in R.

t-test syntax 1:
t.test(x, y = NULL, 
       alternative = c("two.sided", "less", "greater"),
       mu = 0, paired = FALSE, var.equal = FALSE,
       conf.level = 0.95, ...)


t-test syntax 2:
 t.test(formula, data, subset, na.action, ...)


# 1. One sample t-test

One sample t-test is used when we want to compare one group to a hypothetical value.

Suppose a car manufacturer claims that a model has a mpg >= 26. We want to test whether this is true by randomly collect mpg information from 10 cars of the model. Thus, the null and alternative hypotheses are:

H0: mean(mpg)>=26

Ha: mean(mpg) < 26


```{r}
# 10 mpg data points we collected
mpg = c(24.2, 25.3, 23.8, 27.1, 28.0, 25.4, 22.7, 24.6, 25.9, 23.1)

mean(mpg)
```

```{r}
# One sample t-test
t.test(mpg, alternative = "less", mu = 26)
```

Based on the above t-test, the null hypothesis is rejected ( p < 0.05). Thus, the manufacturer's claim is suspicious.

# 2. Two sample t-test

Two sample t-test is used when we want to compare two groups (paired or unpaired).

In this case, we'll use the sleep dataset which shows the effect of two soporific drugs (increase in hours of sleep compared to control) on 10 patients. Our research question is:

Do the two soporific drugs have different effects on extra hours of sleep?

Therefore, the null and alternative hypotheses are defined below:

H0: mean(extra sleep) of drug 1 group = mean(extra sleep) of drug 2 group

Ha: mean(extra sleep) of drug 1 group <> mean(extra sleep) of drug 2 group


```{r}
str(sleep)
```

First, let's draw a box plot to show the extra sleep hours between the two groups.

```{r}
# Draw box plot
plot(extra ~ group, data = sleep, col = c("green","yellow"))
```

From the above box plot, we can find that the mean extra sleep hours seem to be different between the two drug groups. Now, let's formally test the hypothesis.

```{r}
# Select extra sleep hours for two drug groups
extra1 <- sleep$extra[sleep$group == 1]
extra2 <- sleep$extra[sleep$group == 2]

t.test(extra1, extra2, alternative = "two.sided", mu = 0, paired = TRUE)

```

```{r}
# A simpler syntax
t.test(extra1, extra2, paired = TRUE)
```


```{r}
# Use formula
t.test(extra ~ group, data = sleep, paired = TRUE)
```

Whatever type of the specification method you use, you'll get evidance that the The null hypothesis is rejected ( p < 0.01). Thus, the two drugs have significantly different effects on extra hours of sleep.

Pay attention that this is a two sample paired t-test. If you mistakenly use the unpaired t-test, the result will be different. For example,

```{r}
t.test(extra ~ group, data = sleep)
```

