---
title: "Data Exploration - Correlation Analysis"
author: "Cui Zou"
date: "Oct 11, 2019"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage


```{r}
# Clean the environment
rm(list = ls())
```


# 1. Covariance

We can simply use the coviance formula to calculate the covariance between two variables.

```{r}
x <- mtcars$mpg
y <- mtcars$wt
n <- length(x)
x_hat <- mean(x)
y_hat <- mean(y)

sum((x-x_hat)*(y-y_hat))/(n-1)

```

Another way is to define a function to calculate covriance between two variables.

```{r}
# Define a function to calculate covariance
mycov <- function(x,y){
  n <- length(x)
  x_hat <- mean(x)
  y_hat <- mean(y)
  sum((x-x_hat)*(y-y_hat))/(n-1)
}

```

```{r}
# Call the function
cat("covariance =", mycov(mtcars$mpg,mtcars$wt))
```

```{r}
# Another function call
cat("covariance =", mycov(mtcars$hp,mtcars$wt))
```

A simpler way is to directly call the cov() function in the base R system.

```{r}
# Use cov() function to calculate variance
cov(mtcars$mpg,mtcars$wt)
```

```{r}
cov(mtcars$hp,mtcars$wt)
```


# 2. Correlation

Similar to the covariance, we can manually calculate correlation coefficient.

```{r}
# Manually calculate correlation coefficient
cov(mtcars$mpg,mtcars$wt)/(sd(mtcars$mpg)*sd(mtcars$wt))
```

Or we can simply call the cor() function to do the job.

```{r}
# Use cor() function to calculate correlation coefficient
cor(mtcars$mpg,mtcars$wt)
```

```{r}
cor(mtcars$hp,mtcars$wt)
```

# 3. Zero Covariance/Correlation

Covariance and correlation can only measure linear relationship. A zero covariance or correlation does NOT mean that the two variables have no relationship.

In the following example, we create two variables a and b and plot the relationship between these two variables.

```{r}
a <- seq(from = -5, to = 5, by = 0.1)
b <- a^2

plot(a,b)

# Add linear regression line
abline(lm(b~a),col = "red") 
```

From the above plot, we can see that a and b have a curvilinear relationship, b = a^2 + error.

However, as shown below, the covariance and correlation between a and b are very small (close to zero).

```{r}
cov(a,b)
```

```{r}
cor(a,b)
```

# 4. Visualization of a Correlation Matrix

We can use the corrplot package to visualize correlation relationship between variables. This is useful especially when we have many variables.

```{r}
# install.packages("corrplot")
library(corrplot)
```

```{r}
corrplot(cor(mtcars), method="number", tl.cex = 1)
```

```{r}
corrplot(cor(mtcars), method="ellipse", tl.cex = 1)
```
