---
title: "Data Exploration - Dealing with Outliers"
author: "Cui Zou"
date: "Oct 11, 2019"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
&nbsp;


\newpage


```{r}
# Clean the environment
rm(list = ls())
```


# 1. Detect outliers according to boxplot rule

Let's generate 100 random values and manually change three elements as outliers.

```{r}
set.seed(1)
x <- rnorm(100)
# Intentionally add 3 outliers
x <- c(x,-3.5, 3.3, 4.1)

```

```{r}
plot(x)
```

Let's use the boxplot rule to detect outliers.

```{r}
# Create boxplot
bp <- boxplot(x)
```

```{r}
# Print boxplot object
print(bp)
```

We notice that the boxplot object contains an element called "out". That is where the outliers are stored!

```{r}
# Print outliers
bp$out
```



# 2. Detect outliers based on z-score

Again, let's generate 100 random values.

```{r}
# Generate 100 random values
set.seed(1)
x <- rnorm(100)
```
Calculate the mean.
```{r}
mean(x)
```

Calculate standard deviation.

```{r}
sd(x)
```

```{r}
# Intentionally add 3 potential outliers
x <- c(x,-3.5, 3.3, 4.1)

# Calculate mean
mean(x)
```
```{r}
# Calculate standard deviation
sd(x)
```

We notice that the mean and standard deviation have been changed much after the three potential outliers have been added into the dataset.

Now, let's calculate z score and use it to detect potential outliers. All values that have z scores beyond the range of [-3, 3] are potential outliers.

```{r}
# Calculate z-score
z <- (x - mean(x))/sd(x)

# Print outliers
x[which(z > 3 | z < -3)] 
```

We noticed that we can only detect 2 outliers rather than 3. Why?

In the 2nd condition, make sure to have a space between "<" and "-". Or else R may interpret "<-" as assignment rather than comparison.

# 3. Detecting outliers using LOF algorithm

Let's use the famous (Fisher's or Anderson's) iris dataset. The dataset gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.

```{r}
# Use the iris dataset, ignoring the last (5th) column
dat <- iris[-5]

str(dat)
```

Calculate local outlier factor score for each observation. Partition all observations into k = 5 clusters.

```{r}
# Load DMwR package
library(DMwR)

# Call the lofactor() in the DMwR package to calculate local outlier factor score
dat$lof <- lofactor(data = dat, k = 5)

str(dat)

```

We can plot the density of the LOF scores.

```{r}
plot(density(dat$lof))

```

Pick top 5 as outliers.

```{r}
library(magrittr)

outliers <- dat %>% dplyr::arrange(desc(lof)) %>%
  head(5)

print(outliers)
```

Remove outliers from dataset.

```{r}
dat2 <- dplyr::filter(dat,!dat$lof %in% outliers$lof)

str(dat2)
```

Notice that we use the %in% operator to evaluate whether LHS elements are in RHS. For example,

```{r}
## In set or not in set judgement
1:4 %in% 1:2
```

```{r}
!1:4 %in% 1:2
```

