---
title: "Handling NAs"
author: "Cui Zou"
date: "10/13/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
```

**airquality Details**
Daily readings of the following air quality values for May 1, 1973 (a Tuesday) to September 30, 1973.

* Ozone: Mean ozone in parts per billion from 1300 to 1500 hours at Roosevelt Island

* Solar.R: Solar radiation in Langleys in the frequency band 4000--7700 Angstroms from 0800 to 1200 hours at Central Park

* Wind: Average wind speed in miles per hour at 0700 and 1000 hours at LaGuardia Airport

* Temp: Maximum daily temperature in degrees Fahrenheit at La Guardia Airport.

*https://www.rdocumentation.org/packages/datasets/versions/3.6.1/topics/airquality*

```{r}
# Load airquality data
data("airquality")
str(airquality)
```

```{r}
# Find missing values
is.na(airquality)
```

```{r}
# Count missing values
sum(is.na(airquality))
```

```{r}
# Or use summary() function
summary(airquality)

```

```{r}
# Locate the positions with missing values in selected columns
which(is.na(airquality$Ozone))
which(is.na(airquality$Solar.R))
```

```{r}
# Or you can use apply() function to identify all the columns with missing values
list_na <- colnames(airquality)[apply(airquality, 2, anyNA)]
list_na
```

```{r}
# Visualize all the missing values
library(naniar)
vis_miss(airquality)
```

```{r}
# Drop rows with missing values
na.omit(airquality)

```

```{r}
# If dropping all missing values impact your data size dramatically, you may replace those NAs with other values
airquality$Ozone[is.na(airquality$Ozone)] <- mean(airquality$Ozone, na.rm=TRUE)
airquality$Ozone
```


** NHANES example - all variables numerical **
Description
A small data set with non-monotone missing values.

Format
A data frame with 25 observations on the following 4 variables.

* age
Age group (1=20-39, 2=40-59, 3=60+)

* bmi
Body mass index (kg/m**2)

* hyp
Hypertensive (1=no,2=yes)

* chl
Total serum cholesterol (mg/dL)

```{r}
#Loading the mice package
library(mice)
data(nhanes)
str(nhanes)
head(nhanes)

library(lattice)

```

The str function shows us that bmi, hyp and chl has NA values which means missing values. The age variable does not happen to have any missing values. The age values are only 1, 2 and 3 which indicate the age bands 20-39, 40-59 and 60+ respectively. These values are better represented as factors rather than numeric. Let’s convert them:

```{r}
#Convert Age to factor
nhanes$age=as.factor(nhanes$age)
```

```{r}
#understand the missing value pattern
md.pattern(nhanes)
```

1’s and 0’s under each variable represent their presence and missing state respectively. The numbers before the first variable (13,3,1,1,7 here) represent the number of rows. For example, there are 3 cases where chl is missing and all other values are present.There are 5 different missingness patterns. 

```{r}
library(VIM)
#plot the missing values
nhanes_miss = aggr(nhanes, col=mdc(1:2), numbers=TRUE, 
                   sortVars=TRUE, labels=names(nhanes), 
                   cex.axis=.7, gap=3, 
                   ylab=c("Proportion of missingness","Missingness Pattern"))
```

We see that the variables have missing values from 30-40%, so it may not be a good idea to drop all those missing values as our sample size will become much smaller. It also shows the different types of missing patterns and their ratios. 

```{r}
#Imputing missing values using mice
mice_imputes <- mice(nhanes, m=5, maxit = 40)
```

Three parameters for the MICE package: The first is the dataset, the second is the number of times the model should run. I have used the default value of 5 here. This means that I now have 5 imputed datasets. Every dataset was created after a maximum of 40 iterations which is indicated by “maxit” parameter.

```{r}
# Check methods were used for imputing
mice_imputes$method

```

```{r}
# Let's look at our imputed values for chl
mice_imputes$imp$chl
```

According to missing value patterns, chl has 10 missing values with row numbers indicated by the first column. The next five columns show the imputed values because we asked MICE to run 5 times.Now the question is **which dataset to use to fill missing values?**

```{r}
# Plotting and comparing values with xyplot()
xyplot(mice_imputes, bmi ~ chl | .imp, pch = 20, cex = 1.4)
```

The blue ones are the observed data and red ones are imputed data. The red points should ideally be similar to the blue ones so that the imputed values are similar. 

```{r}
# Imputed dataset
# Since the 4th dataset gave us the best fitting, I will impute the missing values from it.
Imputed_data=complete(mice_imputes,5)
summary(Imputed_data)
```







